TRIK Studio: Technical Introduction

### 2017-02-10 2017-trikStudioEnglish/trikStudioEnglish.tex
Minor fix

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to comprehend, so it suits the purposes of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: a program controlling a robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where a program consists of a number of "`black boxes"', connected with data channels. Each such "`box"' (we will call it \emph{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call \emph{tokens} data transferring through channels). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sensor data to its output. Many researchers  note usability of visual data-flow languages comparing to control-flow ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It 	emerged as a further evolution of QReal:Robots~\cite{terekhov2013robots} project, developed at Software Engineering chair of Saint-Petersburg State University. It supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (more simple control-flow and more complex data-flow language) or one of a number of textual ones. For Lego NXT the programmer can choose from NXT OSEK C and Russian version of C (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual machine's byte code, is supported.

@@

\begin{itemize}
    \item The first group is for blocks implementing basic algorithmic expressions, like a start and an end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for parallel execution and for working with concurrent tasks (e.g. for merging and interrupting tasks), subprogram call block and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, there are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that "`freeze"' the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator game pad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text and images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{TS_CF_Labyrinth_Diagram.png}
    \caption{The program of traversing the maze using right-hand rule. The first four blocks initialize the program: set the wall proximity variable, turn the robot motors on, after that the robot moves for one second to enter the maze. The second four blocks define the main control loop. Each ten milliseconds the robot checks using its infrared sensors if it can move forward or right (the \texttt{sensorA2 <= threshold \&\& sensorA1 <= threshold} condition). If it can move forward, the movement will be performed using proportional control over the right sensor data (the "`Regulator"' subprogram). If there are walls in front of the robot and to the right, it turn 90 degrees left (the "`Turn left"' subprogram)}.
    \label{image:TS_CF_Example}
\end{figure*}

@@

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. The editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}. Language metamodel was defined using QReal's metaeditor, and a module implementing the editor was automatically generated using QReal's tools (as some other technologies like~\cite{bryksin2011ubiq}). Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE "`inherits"' all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~(\cite{osechkina2010gestures,osechkina2012multistroke}), copy-paste and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent only about three man-days working on it. We believe that it acknowledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

To make transition to this "`advanced"' language simpler it maintains some kind of conceptual compatibility with the control-flow language. Blocks can also be organized in a chain with explicit passing of control, to make it possible most blocks have a special activation port that ignores all input data and just executes the block like it was in the control-flow case. So users can write programs like they used to and move on to using smaller number of blocks as they gain more experienced.

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's "`colony"' architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not provided here. Another ideas on this matter could be found, for example, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only a Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Furthermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not "`know"' about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extension modules supporting new languages, robotic kits and tools.

@@

An interesting consequence of this approach is that users can change diagrams during interpretation and changes will be applied "`on the fly"'. It is a convenient behavior, for example, for selecting parameters of a PID controller in a program. Meanwhile, the execution of unchanged diagram parts is optimized: block implementation objects are cached, the same is true for abstract syntax trees and type inference information for textual programs. Thus, if the program remains unchanged while being interpreted, subsequent control flow to the same parts of the diagram will employ already created objects. Diagram validation is also performed "`on the fly"'. If some incorrect fragment is found, an error message is shown. It could be seen as a drawback since users get their error messages only when such errors are reached during interpretation, but this behavior is typical for all widely used interpreted textual languages.

@@

One of the most interesting and demanded TRIK Studio features is generation of well-readable textual code from visual diagrams. At the moment of writing several languages are supported for code generation: C, JavaScript, F\#, Pascal, Russian version of C~\cite{тереховотечественные} and byte code for Lego EV3 virtual machine. This section presents only code generators for control-flow programs.

@@

\section{Textual language parser and interpreter}
\label{chapter:parser}
TRIK Studio uses visual languages for robots programming, but arithmetic expressions, intrinsic function calls and so on are better represented with textual strings (in fact, NXT-G tries to use visual blocks even for arithmetic expressions, representing them as syntax trees, it is very inconvenient). As a textual part of a language for both visual languages TRIK Studio uses a subset of Lua 5.3\footnote{https://www.lua.org/ (accessed: 07.02.2017)}, customized for our needs. The Decision to implement our own parser and interpreter of a textual language was made under these considerations:

@@

QReal parser combinator library supports recursive-descent parsers which are able to parse a subset of LL(1) grammars. FOLLOW($\alpha$) set is not calculated, so it limits the expressiveness of resulting grammars, but we were still able to use Lua 5.3 grammar almost as it is specified, with only minor modifications related to factorization and left recursion elimination, which shall be done anyway for LL parsers. For arithmetic expressions Precedence Climbing\footnote{http://www.engr.mun.ca/\~theo/Misc/exp\_parsing.htm (accessed: 07.02.2017)} algorithm was used, and it also required some minor alterations in Lua grammar. Custom modifications were made mainly on lexer level to allow, for example, to use '!=' for inequality in addition to '\verb|~=|' used in Lua. Here is a quick example\footnote{for full specification of our parser with many irrelevant technical details see https://github.com/qreal/qreal/tree/master/qrtext/src/lua} of production written in C++ with our library ("`statement is ';' or a list of expressions, optionally followed by '=' and other list of expressions"'):
\begin{verbatim}
	// stat ::= ‘;’ | explist [‘=’ explist]
	stat = (-LuaTokenTypes::semicolon 
	    | (explist 
			   & ~(-LuaTokenTypes::equals & explist)))
\end{verbatim}

\verb|LuaTokenTypes::semicolon| is a token corresponding to ';', operator '-' creates a simple parser that can parse only semicolons and excludes semicolons from the AST, "`explist"` is a reference parser object, like "`stat"', defined elsewhere, '\&' combines two parsers into a parser which accepts concatenation of their corresponding strings, '|' combines two parsers into a parser that accepts alternative, '\verb|~|' creates an optional parser from its argument, which shall also be a parser. There are also operators for adding semantic actions to productions and assigning parser a name for debug purposes. When all parsers are combined in such a way, it is enough to call 'parse()' method of a resulting parser object, giving it a token stream. The parser combinator library was used for another language in QReal~\cite{tikhonova2015generation}, so it is general enough to support not only Lua.

@@

The simulation environment has two implementations of robot behavior models: "`ideal"' and "`realistic"'. Within the ideal model all friction forces (for floor and walls) are ignored, and having constant motor powers the robot continues to move without any acceleration or slowdown. Any, even slightest collision with a wall will stop the robot. The realistic model takes care of thrust and friction force, when crashing into a wall the robot tries to behave like the real device. Users also have an ability to add Gaussian noise to sensor data and motor impulses. Changing between these models is done using checkboxs in the simulation settings window and could be done even during the execution of a program.

@@

Each event is specified by its XML node in a program. An event can be specified in a \textit{canonical} form, or in a form of a \textit{constraint}. An event in a canonical form is a previously described triple $(id_i, c_i, T_i)$. An event in a constraint form is a triple $(id_j, c_j, message_j)$. A constraint is interpreted as an event $(id_i, $$\neg$$c_i, [ fail(message_j) ])$, where $fail(message_j)$ is a trigger that stops simulation and reports a $message_j$ error message. In other words, a constraint in an event that is raised when given condition is violated and reports this violation to a user. Events in a form of constraints are added to a language for pragmatical reasons only, because it is much more convenient to specify conditions like "`robot $x$ shall not leave area $a$"' or "`robot $x$ shall have a set of sensors $s$ connected"' as constraints instead of events. Time limits are a special case of such constraints. A time limit shall be specified for each constraint definition program so the constraints checker will not run the simulation indefinite amount of time if all constraints are satisfied but success event is never triggered. The checker verifies that a time limit is set before execution of a program, and if not, considers it as a semantic error.

Let's briefly describe predicates,functional symbols and elementary triggers used in the language. Predicate symbols are be divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an "`item $x$ is located inside area $y$"' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other predicates that can be expressed by already described predicates, they are introduced for pragmatical reasons only.
\end{itemize}

@@

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)}. Without such a checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch "`headless"' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker\footnote{https://www.docker.com/ (accessed: 07.02.2017)} container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users but error reporter output does gets displayed, so students can guess what went wrong (well, theoretically, but to our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).

The checker can also send a visual representation of a field with a trace of a robot to a client. A trace consists of points of a robot trajectory and values reported by its sensors, and this trace can be played back by a web application described in~\cite{zakharov2016web}. This web application also allows to actually create solutions for almost all tasks from the course right in the browser, form a correct TRIK Studio save file and run it on a server on the checker, playing back the result, so a student perceives this as "`TRIK Studio in a browser"'. This application works standalone, but is not integrated into the Stepik course yet due to technical difficulties related to the Stepik infrastructure.

@@

%\bibliographystyle{utf8gost705u}
\bibliographystyle{ieeetran}
\bibliography{trikStudioEnglish}

>>>>

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to comprehend, so it suits the purposes of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: a program controlling a robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where a program consists of a number of ``black boxes'', connected with data channels. Each such ``box'' (we will call it \emph{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call \emph{tokens} data transferring through channels). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sensor data to its output. Many researchers note usability of visual data-flow languages comparing to a textual ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It emerged as a further evolution of QReal:Robots~\cite{terekhov2013robots} project, developed at Software Engineering chair of Saint-Petersburg State University. It supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (more simple control-flow and more complex data-flow language) or one of a number of textual ones. For Lego NXT the programmer can choose between NXT OSEK C and Russian version of C (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual machine's byte code, is supported.

@@

\begin{itemize}
    \item The first group is for blocks implementing basic algorithmic expressions, like a start and an end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for parallel execution and for working with concurrent tasks (e.g. for merging and interrupting tasks), subprogram call block and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, there are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that ``freeze'' the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator game pad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text and images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{TS_CF_Labyrinth_Diagram.png}
    \caption{The program of traversing the maze using right-hand rule. The first four blocks initialize the program: set the wall proximity variable, turn the robot motors on, after that the robot moves for one second to enter the maze. The second four blocks define the main control loop. Each ten milliseconds the robot checks using its infrared sensors if it can move forward or right (the \texttt{sensorA2 <= threshold \&\& sensorA1 <= threshold} condition). If it can move forward, the movement will be performed using proportional control over the right sensor data (the ``Regulator'' subprogram). If there are walls in front of the robot and to the right, it turn 90 degrees left (the ``Turn left'' subprogram)}.
    \label{image:TS_CF_Example}
\end{figure*}

@@

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. The editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}. Language metamodel was defined using QReal's metaeditor, and a module implementing the editor was automatically generated using QReal's tools (as some other technologies like~\cite{bryksin2011ubiq}). Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE ``inherits'' all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~(\cite{osechkina2010gestures,osechkina2012multistroke}), copy-paste and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent only about three man-days working on it. We believe that it acknowledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

To make transition to this ``advanced'' language simpler it maintains some kind of conceptual compatibility with the control-flow language. Blocks can also be organized in a chain with explicit passing of control, to make it possible most blocks have a special activation port that ignores all input data and just executes the block like it was in the control-flow case. So users can write programs like they used to and move on to using smaller number of blocks as they gain more experienced.

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's ``colony'' architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not provided here. Another ideas on this matter could be found, for example, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only a Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Furthermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not ``know'' about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extension modules supporting new languages, robotic kits and tools.

@@

An interesting consequence of this approach is that users can change diagrams during interpretation and changes will be applied ``on the fly''. It is a convenient behavior, for example, for selecting parameters of a PID controller in a program. Meanwhile, the execution of unchanged diagram parts is optimized: block implementation objects are cached, the same is true for abstract syntax trees and type inference information for textual programs. Thus, if the program remains unchanged while being interpreted, subsequent control flow to the same parts of the diagram will employ already created objects. Diagram validation is also performed ``on the fly''. If some incorrect fragment is found, an error message is shown. It could be seen as a drawback since users get their error messages only when such errors are reached during interpretation, but this behavior is typical for all widely used interpreted textual languages.

@@

One of the most interesting and demanded TRIK Studio features is generation of well-readable textual code from visual diagrams. Several languages are supported for code generation: C, JavaScript, F\#, Pascal, Russian version of C~\cite{тереховотечественные} and byte code for Lego EV3 virtual machine. This section presents only code generators for control-flow programs.

@@

\section{Textual language parser and interpreter}
\label{chapter:parser}
TRIK Studio uses visual languages for robots programming, but arithmetic expressions, intrinsic function calls and so on are better represented with textual strings (in fact, NXT-G tries to use visual blocks even for arithmetic expressions, representing them as syntax trees, it is very inconvenient). As a textual part of a language for both visual languages TRIK Studio uses a subset of Lua 5.3\footnote{https://www.lua.org/ (accessed: 07.02.2017)}, customized for our needs. The decision to implement our own parser and interpreter of a textual language was made under these considerations:

@@

QReal parser combinator library supports recursive-descent parsers which are able to parse a subset of LL(1) grammars. FOLLOW($\alpha$) set is not calculated, so it limits the expressiveness of resulting grammars, but we were still able to use Lua 5.3 grammar almost as it is specified, with only minor modifications related to factorization and left recursion elimination, which shall be done anyway for LL parsers. For arithmetic expressions Precedence Climbing\footnote{http://www.engr.mun.ca/\~theo/Misc/exp\_parsing.htm (accessed: 07.02.2017)} algorithm was used, and it also required some minor alterations in Lua grammar. Custom modifications were made mainly on lexer level to allow, for example, to use '!=' for inequality in addition to '\verb|~=|' used in Lua. Here is a quick example\footnote{for full specification of our parser with many irrelevant technical details see https://github.com/qreal/qreal/tree/master/qrtext/src/lua} of production written in C++ with our library (``statement is ';' or a list of expressions, optionally followed by '=' and other list of expressions''):
\begin{verbatim}
	// stat ::= ‘;’ | explist [‘=’ explist]
	stat = (-LuaTokenTypes::semicolon 
	    | (explist 
			   & ~(-LuaTokenTypes::equals & explist)))
\end{verbatim}

\verb|LuaTokenTypes::semicolon| is a token corresponding to ';', operator '-' creates a simple parser that can parse only semicolons and excludes semicolons from the AST, ``explist'' is a reference parser object, like ``stat'', defined elsewhere, '\&' combines two parsers into a parser which accepts concatenation of their corresponding strings, '|' combines two parsers into a parser that accepts alternative, '\verb|~|' creates an optional parser from its argument, which shall also be a parser. There are also operators for adding semantic actions to productions and assigning parser a name for debug purposes. When all parsers are combined in such a way, it is enough to call 'parse()' method of a resulting parser object, giving it a token stream. The parser combinator library was used for another language in QReal~\cite{tikhonova2015generation}, so it is general enough to support not only Lua.

@@

The simulation environment has two implementations of robot behavior models: ``ideal'' and ``realistic''. Within the ideal model all friction forces (for floor and walls) are ignored, and having constant motor powers the robot continues to move without any acceleration or slowdown. Any, even slightest collision with a wall will stop the robot. The realistic model takes care of thrust and friction force, when crashing into a wall the robot tries to behave like the real device. Users also have an ability to add Gaussian noise to sensor data and motor impulses. Changing between these models is done using checkboxs in the simulation settings window and could be done even during the execution of a program.

@@

Each event is specified by its XML node in a program. An event can be specified in a \textit{canonical} form, or in a form of a \textit{constraint}. An event in a canonical form is a previously described triple $(id_i, c_i, T_i)$. An event in a constraint form is a triple $(id_j, c_j, message_j)$. A constraint is interpreted as an event $(id_i, $$\neg$$c_i, [ fail(message_j) ])$, where $fail(message_j)$ is a trigger that stops simulation and reports a $message_j$ error message. In other words, a constraint in an event that is raised when given condition is violated and reports this violation to a user. Events in a form of constraints are added to a language for pragmatical reasons only, because it is much more convenient to specify conditions like ``robot $x$ shall not leave area $a$'' or ``robot $x$ shall have a set of sensors $s$ connected'' as constraints instead of events. Time limits are a special case of such constraints. A time limit shall be specified for each constraint definition program so the constraints checker will not run the simulation indefinite amount of time if all constraints are satisfied but success event is never triggered. The checker verifies that a time limit is set before execution of a program, and if not, considers it as a semantic error.

Let's briefly describe predicates,functional symbols and elementary triggers used in the language. Predicate symbols are be divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an ``item $x$ is located inside area $y$'' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other predicates that can be expressed by already described predicates, they are introduced for pragmatical reasons only.
\end{itemize}

@@

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)}. Without such a checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch ``headless'' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker\footnote{https://www.docker.com/ (accessed: 07.02.2017)} container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users but error reporter output does gets displayed, so students can guess what went wrong (well, theoretically, but to our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).

The checker can also send a visual representation of a field with a trace of a robot to a client. A trace consists of points of a robot trajectory and values reported by its sensors, and this trace can be played back by a web application described in~\cite{zakharov2016web}. This web application also allows to actually create solutions for almost all tasks from the course right in the browser, form a correct TRIK Studio save file and run it on a server on the checker, playing back the result, so a student perceives this as ``TRIK Studio in a browser''. This application works standalone, but is not integrated into the Stepik course yet due to technical difficulties related to the Stepik infrastructure.

@@

\bibliographystyle{IEEEtran}
\bibliography{trikStudioEnglish}
